---
title: "GDP per Capita vs Fertility Rate"
author: "Sumanth, Amar, Davis, James"
format: html
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

In this project, we will explore the relationship between **Total Fertility Rate** and **GDP per Capita** using data from Gapminder. These variables were chosen because we believe economic development influences population growth.

### Variables

-   **Total Fertility Rate (TFR):** Average number of children born to a woman over her lifetime.
    -   Source: [Gapminder](https://www.gapminder.org/data/), drawn from the World Bank data.
-   **GDP per Capita:** The economic output per person, adjusted for inflation and population. In 2017 dollars.
    -   Source: [Gapminder](https://www.gapminder.org/data/), drawn from the World Bank data.

## Hypothesis

> As GDP per capita increases, Total Fertility Rate increases.

From our research, we found many sources that imply that economically developed nations tend to have greater access to education, better access to reproductive healthcare and contraception, and an increased cost of raising children. We hypothesized that these factors would all lead to a decrease in the Total Fertility Rate.

According to the Our World in Data [Fertility vs Population Rate (2019)](https://ourworldindata.org/grapher/children-per-woman-fertility-rate-vs-level-of-prosperity), this trend is observed globally as fertility transitions follow a downwards trend with gdp.

## Data Cleaning and Preparation

```{r}
library(tidyverse)
library(knitr)
library(ggplot2)
library(gganimate)
library(transformr)
library(gifski)
library(kableExtra)
```

### Load data files

```{r}
tfr <- read_csv("sp_dyn_tfrt_in.csv") 
gdp <- read_csv("gdp_pcap.csv")
```

### Getting rid of k

The GDP data has some values with a "k" suffix, which indicates thousands. We need to convert these values to numeric.

```{r}
gdp_clean <- gdp |>
  mutate(across(-country, ~ case_when(
    str_detect(., "k") ~ as.numeric(str_remove(., "k")) * 1000,
    TRUE ~ as.numeric(.)
  )))
```

### Pivoting to long format

With this, we can easily join the two datasets on country and year.

```{r}
tfr_long <- tfr |>
  pivot_longer(cols = -country, names_to = "year", values_to = "tfr") |>
  mutate(year = as.integer(year))

gdp_long <- gdp_clean |>
  pivot_longer(cols = -country, names_to = "year", values_to = "gdp_per_capita") |>
  mutate(year = as.integer(year))



```

### Join the datasets on country and year

This will be the dataset we will work with for the rest of the project.

```{r}
data_clean <- inner_join(tfr_long, gdp_long, by = c("country", "year")) |>
  filter(!is.na(tfr), !is.na(gdp_per_capita))


data_clean |>
  head() |>
  kable()
```

### Data Visualizations 

This is the relationship between Average Fertility Rate vs GDP per Capita by Country.
```{r}
agg_data <- data_clean |>
  group_by(country) |>
  summarise(
    avg_tfr = mean(tfr, na.rm = TRUE),
    avg_gdp = mean(gdp_per_capita, na.rm = TRUE)
  )

ggplot(agg_data, aes(x = avg_gdp, y = avg_tfr)) +
  geom_point(color = "steelblue", alpha = 0.7, size = 2) +
  labs(
    title = "Average Total Fertility Rate vs. GDP per Capita",
    subtitle = "Aggregated by Country (2017 USD)",
    x = "Average GDP per Capita (2017 USD)",
    y = "Average Total Fertility Rate"
  )
```

```{r}
agg_data <- data_clean |>
  group_by(country) |>
  summarise(
    avg_tfr = mean(tfr, na.rm = TRUE),
    avg_gdp = mean(gdp_per_capita, na.rm = TRUE)
  )

ggplot(agg_data, aes(x = log(avg_gdp), y = log(avg_tfr))) +
  geom_point(color = "steelblue", alpha = 0.7, size = 2) +
  labs(
    title = "Log Average Total Fertility Rate vs. GDP per Capita",
    subtitle = "Aggregated by Country (2017 USD)",
    x = "Log Average GDP per Capita (2017 USD)",
    y = "Log Average Total Fertility Rate"
  )
```

This is the relationship between Average Fertility Rate vs GDP per Capita by Country over time using an animated plot.

```{r}
ggplot(data_clean, aes(x = log(gdp_per_capita), y = log(tfr))) +
  geom_point(aes(group = country), alpha = 0.6, size = 1.5, color = "steelblue") +
  labs(
    title = "Log Fertility vs GDP per Capita Over Time: {closest_state}",
    x = "Log GDP per Capita (2017 USD)",
    y = "Log Total Fertility Rate"
  ) +
  theme_light(base_size = 12) +
  transition_states(year, transition_length = 1, state_length = 1) +
  ease_aes('linear')
```


2.2 Linear Regression
```{r}
ln_model <- lm(log(avg_tfr) ~ log(avg_gdp), data = agg_data)
broom::tidy(ln_model) |>
  kable()

```
Intercept: If the average gdp per capita is 1 dollar in 2017 dollars, then the average fertility rate 54 children which is not meaningful.

Slope: A 2-fold increase in average gdp per capita decreases the predicted average value of average fertility rate by a factor of 0.805.


2.3 Model fit

# Table

```{r}
augmented_data <- broom::augment(ln_model)

tibble(
    variance = var(ln_model$model[[1]]),
    variance_fitted = var(fitted(ln_model)),
    variance_res = summary(ln_model)$sigma^2,
    rsquared = summary(ln_model)$r.squared 
  ) |>
  summarize(Variance = variance,
            'Fitted Values' = variance_fitted,
            'Residual Variance' = variance_res,
            '$R^2$' = rsquared)|>
  kable(booktabs = TRUE, caption = "Model Variance Summary")|>
  kable_styling(latex_options = "striped")



```

Discuss the proportion of the variability in the response values that was accounted for by your regression model. What does this suggest about the “quality” of your model?


With an $R^2$ value of .5879, our model explains 58.8% of the variance in average total fertility rate across countries. This indicates that average GDP per capita is a decent predictor for explaining average total fertility rate. While 58% of the data is explained by our model, around 41% is still left unexplained. To increase our $R^2$, we can possibly add more predictors to our model or choose different predictors that explain more variation than average GDP per capita.








