
---
title: "GDP per Capita vs Fertility Rate"
author: "Sumanth, Amar, Davis, James"
format: 
  html:
    embed-resources: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

In this project, we will explore the relationship between **Total Fertility Rate** and **GDP per Capita** using data from Gapminder. These variables were chosen because we believe economic development influences population growth.

### Variables

-   **Total Fertility Rate (TFR):** Average number of children born to a woman over her lifetime.
    -   Source: [Gapminder](https://www.gapminder.org/data/), drawn from the World Bank data.
-   **GDP per Capita:** The economic output per person, adjusted for inflation and population. In 2017 dollars.
    -   Source: [Gapminder](https://www.gapminder.org/data/), drawn from the World Bank data.

## Hypothesis

> As GDP per capita increases, Total Fertility Rate decreases.

From our research, we found many sources that imply that economically developed nations tend to have greater access to education, better access to reproductive healthcare and contraception, and an increased cost of raising children. We hypothesized that these factors would all lead to a decrease in the Total Fertility Rate.

According to the Our World in Data [Fertility vs Population Rate (2019)](https://ourworldindata.org/grapher/children-per-woman-fertility-rate-vs-level-of-prosperity), this trend is observed globally as fertility transitions follow a downwards trend with GDP.

## Data Cleaning and Preparation

We began by standardizing the GDP, which included some values reported in thousands (denoted with a "k"). These values were converted to numerical form by multiplying them by 1,000 and dropping the "k". After aligning the year columns and ensuring both datasets were in long format, we joined them based on country and year. Rows with missing values for either GDP or fertility rate were removed. The result was a clean dataset containing matched pairs of GDP per capita and fertility rates across countries and years.

The unit of observation in our analysis is a country-year pair. We aggregated this data at the country level by calculating the average fertility rate and average GDP per capita for each country.

```{r}
library(tidyverse)
library(knitr)
library(ggplot2)
library(gganimate)
library(transformr)
library(gifski)
library(kableExtra)
```

```{r Load Data Files}
tfr <- read_csv("sp_dyn_tfrt_in.csv") 
gdp <- read_csv("gdp_pcap.csv")
```

```{r clean_gdp_values}
gdp_clean <- gdp |>
  mutate(across(-country, ~ case_when(
    str_detect(., "k") ~ as.numeric(str_remove(., "k")) * 1000,
    TRUE ~ as.numeric(.)
  )))
```

```{r pivot_to_long_format}
tfr_long <- tfr |>
  pivot_longer(cols = -country, names_to = "year", values_to = "tfr") |>
  mutate(year = as.integer(year))

gdp_long <- gdp_clean |>
  pivot_longer(cols = -country, names_to = "year", values_to = "gdp_per_capita") |>
  mutate(year = as.integer(year))
```

```{r join_datasets_country_year}
data_clean <- inner_join(tfr_long, gdp_long, by = c("country", "year")) |>
  filter(!is.na(tfr), !is.na(gdp_per_capita))
```

## Data Visualizations

Figure 1 illustrates the overall relationship between GDP per capita and total fertility rate, aggregated by country:

```{r}
agg_data <- data_clean |>
  group_by(country) |>
  summarise(
    avg_tfr = mean(tfr, na.rm = TRUE),
    avg_gdp = mean(gdp_per_capita, na.rm = TRUE)
  )

ggplot(agg_data, aes(x = avg_gdp, y = avg_tfr)) +
  geom_point(color = "steelblue", alpha = 0.7, size = 2) +
  labs(
    title = "Average Total Fertility Rate vs. GDP per Capita",
    subtitle = "Aggregated by Country (2017 USD)",
    x = "Average GDP per Capita (2017 USD)",
    y = "Average Total Fertility Rate"
  )
```

In this raw scale plot, we observe a clear negative association, countries with lower GDP per capita often have higher fertility rates. However, the relationship becomes more interpretable and linear when both axes are transformed using logarithms, as shown in Figure 2:

```{r}
ggplot(agg_data, aes(x = log(avg_gdp), y = log(avg_tfr))) +
  geom_point(color = "steelblue", alpha = 0.7, size = 2) +
  labs(
    title = "Log Average Total Fertility Rate vs. GDP per Capita",
    subtitle = "Aggregated by Country (2017 USD)",
    x = "Log Average GDP per Capita (2017 USD)",
    y = "Log Average Total Fertility Rate"
  )
```

This transformation reveals a stronger, more consistent downward trend, suggesting a potential linear relationship on the log scale.

To further understand how this relationship has changed over time, we visualized the data as an animated graphic showing how fertility and GDP move year by year. As seen in the animation, countries’ positions shift notably: while the early years show a scattered and less defined relationship, over time the points gradually shift downward into a linear shape. This movement indicates a collective decline in fertility rates globally, with wealthier nations going lower on the fertility axis. The downward slope becomes more dramatic in more recent years, emphasizing the strengthening inverse relationship between economic development and fertility.

```{r}
ggplot(data_clean, aes(x = log(gdp_per_capita), y = log(tfr))) +
  geom_point(aes(group = country), alpha = 0.6, size = 1.5, color = "steelblue") +
  labs(
    title = "Log Fertility vs GDP per Capita Over Time: {closest_state}",
    x = "Log GDP per Capita (2017 USD)",
    y = "Log Total Fertility Rate"
  ) +
  theme_light(base_size = 12) +
  transition_states(year, transition_length = 1, state_length = 1) +
  ease_aes('linear')
```

## Linear Regression

```{r}
ln_model <- lm(log(avg_tfr) ~ log(avg_gdp), data = agg_data)

broom::tidy(ln_model) |>
  kable()
```

The intercept suggests that if the average GDP per capita were \$1, the predicted fertility rate would be approximately 54 children per woman. 

The slope of the regression model indicates that for every doubling in average GDP per capita, the predicted average total fertility rate decreases by a factor of 0.805. This suggests a negative relationship between economic prosperity and fertility: as countries become wealthier on average, they tend to have lower fertility rates. This aligns with our hypothesis.

## Model Fit

```{r}
augmented_data <- broom::augment(ln_model)

tibble(
  variance = var(ln_model$model[[1]]),
  variance_fitted = var(fitted(ln_model)),
  variance_res = summary(ln_model)$sigma^2,
  rsquared = summary(ln_model)$r.squared 
) |>
  summarize(
    Variance = variance,
    `Fitted Values` = variance_fitted,
    `Residual Variance` = variance_res,
    `$R^2$` = rsquared
  ) |>
  kable(booktabs = TRUE, caption = "Model Variance Summary") |>
  kable_styling(latex_options = "striped")
```

With an $R^2$ value of 0.5879, our model explains 58.8% of the variance in average total fertility rate across countries. This indicates that average GDP per capita is a decent predictor for explaining average total fertility rate. While 58% of the data is explained by our model, around 41% is still left unexplained. To increase our $R^2$, we can possibly add more predictors to our model or choose different predictors that explain more variation than average GDP per capita.

## Cross Validation

### 10-Fold Cross Validation

```{r}
library(purrr)

set.seed(42)

k <- 10

agg_data_folds <- agg_data |>
  mutate(fold = sample(rep(1:k, length.out = n())))

get_fold_r2 <- function(fold_number, data) {
  train_data <- filter(data, fold != fold_number)
  test_data <- filter(data, fold == fold_number)
  
  model <- lm(log(avg_tfr) ~ log(avg_gdp), data = train_data)
  preds <- predict(model, newdata = test_data)
  
  r2 <- var(preds, na.rm = TRUE) / var(log(test_data$avg_tfr), na.rm = TRUE)
  return(r2)
}

r2_values <- map_dbl(1:k, ~get_fold_r2(.x, agg_data_folds))

r2_df <- tibble(fold = 1:k, r2 = r2_values)

r2_df |>
  kable(booktabs = TRUE, caption = "R² Values from 10-Fold Cross Validation") |>
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```

### Plotting the R² Distribution

```{r}
ggplot(r2_df, aes(x = r2)) +
  geom_histogram(
    binwidth = 0.05, 
    fill = "steelblue", 
    color = "black", 
    alpha = 0.8,
    boundary = 0, 
    closed = "left"
  ) +
  labs(
    title = "Distribution of R² Values from 10-Fold Cross Validation",
    subtitle = paste0("Mean R² = ", round(mean(r2_df$r2), 3)),
    x = expression(R^2~"Value"),
    y = "Number of Folds"
  ) +
  theme_minimal(base_size = 13)
```

The results of the 10-fold cross-validation show that the model's $R^2$ values range from approximately 0.369 to 0.910, with an average $R^2$ of about 0.594. This is consistent with the $R^2$ from the full model (0.588), suggesting that the model generalizes well to unseen data and is not overfitting. There is some variation in $R^2$ across folds, likely due to the small sample size and potential variability in GDP and fertility values between countries. Overall, this goes to show that GDP per capita is a decent predictor of average total fertility rate, but there is still room for improvement in the model.

## Conclusion

This analysis provides compelling evidence for a negative association between economic prosperity and fertility rates. As GDP per capita increases, fertility rates tend to decline. The regression results are consistent across visualizations and validated by cross-validation, suggesting that GDP per capita is a meaningful, though not exclusive, predictor of fertility behavior across nations. These findings support existing global demographic theories.

Future research could explore additional factors such as education levels, healthcare access, or cultural variables to improve explanation. 

## References
-   **Total Fertility Rate (TFR):**.
    -   Source: [Gapminder](https://www.gapminder.org/data/), drawn from the World Bank data.
-   **GDP per Capita:** .
    -   Source: [Gapminder](https://www.gapminder.org/data/), drawn from the World Bank data.
- **Fertility vs Population Rate (2019):**
    - Our World in Data [Fertility vs Population Rate (2019)](https://ourworldindata.org/grapher/children-per-woman-fertility-rate-vs-level-of-prosperity)

All code for reproducibility can be found at 
[Stat 331 Project](https://github.com/sumanth2122/stat331-project.git)
    
